---
title: "The DietView data model and input contract"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The DietView data model and input contract}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

DietView is built on a single, deliberately simple contract: **one row is one
prey record in one stomach**. Everything else — taxonomy, metrics, figures,
dashboards — is derived from that. This article explains the contract, the role
system, and how to map *any* survey onto it.

```{r setup}
library(DietView)
```

## One row per prey record

Column names are free. You never rename your data; instead you declare, once,
which column plays which **role**. Here is the bundled base table:

```{r data}
dat <- read.csv(
  system.file("extdata", "dietview_example.csv", package = "DietView"),
  na.strings = c("NA", "")
)
str(dat)
```

Note what is *not* here: there are no `kingdom .. species` columns. DietView adds
those itself (see the taxonomy article).

## Declaring roles with `dv_spec()`

```{r spec}
spec <- dv_spec(
  predator       = "predator_species_common_name",
  predator_latin = "predator_species_latin_name",
  stomach        = "stomach_id",
  prey_id        = "aphia_id",       # WoRMS AphiaID (preferred key)
  prey_name      = "verified_name",  # fallback key
  weight         = "prey_weight",
  digestion      = "digestion_level_id",
  covariates     = c("period", "size_class", "size_bin", "region")
)
```

The roles and their meaning:

| Role | Required | Purpose |
|---|---|---|
| `predator` | yes | one predator per row |
| `stomach` | yes | unit of occurrence; drives stomach counts |
| `prey_id` / `prey_name` | one of them | prey key(s) for taxonomy |
| `weight` | no | the weight currency (`%W`) |
| `count` | no | per-record count for `%N` |
| `digestion` | no | enables the digestion diagnostic |
| `covariates` | no | any stratifiers |
| `predator_latin` | no | auto-fetch ID photos |

## Validating a table against a spec

`dv_validate()` fails fast on hard errors (a missing required column, no prey
key, an absent covariate) and only warns on soft gaps (no weight → `%W` empty; no
digestion → diagnostic disabled). It returns the data invisibly, so it composes
in a pipeline.

```{r validate}
invisible(dv_validate(dat, spec))
```

## Mapping your own survey

Because nothing is hard-coded, adapting DietView to a new dataset is entirely a
matter of writing a different `dv_spec()`. If your columns were, say,
`sp`, `stom`, `aphia`, `wt_g`, and `yr`, you would simply write:

```{r own, eval = FALSE}
spec <- dv_spec(
  predator = "sp", stomach = "stom", prey_id = "aphia",
  weight = "wt_g", covariates = "yr"
)
```

Everything downstream — `dv_prepare()`, `dv_build_tree()`, `dv_sunburst()`,
`dv_build_html()` — reads the spec, never the literal names.
