---
title: "The digestion identification bias"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The digestion identification bias}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5,
                      warning = FALSE, message = FALSE)
```

Diet identification is not uniform: heavily digested prey can only be resolved to
coarse ranks. If that bias is hidden, a diet description silently over-represents
whatever is easy to identify. `dv_digestion_diagnostic()` makes it explicit.

```{r setup}
library(DietView)

dat <- read.csv(system.file("extdata", "dietview_example.csv", package = "DietView"),
                na.strings = c("NA", ""))
lin <- read.csv(system.file("extdata", "dietview_example_lineage.csv", package = "DietView"),
                na.strings = c("NA", ""))
spec <- dv_spec(predator = "predator_species_common_name", stomach = "stomach_id",
                prey_id = "aphia_id", prey_name = "verified_name",
                weight = "prey_weight", digestion = "digestion_level_id",
                covariates = c("period"))
prep <- dv_prepare(dat, spec, taxonomy = "table", lineage = lin)
```

## The diagnostic

For each digestion level, the share of records (and of weight) whose **finest
resolved rank** is Kingdom, Phylum, ..., Species, or Unresolved:

```{r diag}
diag <- dv_digestion_diagnostic(prep)
head(diag, 12)
```

## Reading the bias

As the digestion level increases, mass shifts away from Species toward the coarse
ranks and "Unresolved". A dependency-free stacked bar makes the shift visible:

```{r plot}
m <- xtabs(p_records ~ finest_rank + digestion, data = diag)
barplot(m, col = grDevices::hcl.colors(nrow(m), "Blues"),
        legend.text = TRUE,
        args.legend = list(x = "topright", cex = 0.7, bg = "white"),
        xlab = "digestion level", ylab = "% of records")
```

## Splitting by a covariate

Pass `by =` to see whether the bias differs across, say, periods:

```{r by}
head(dv_digestion_diagnostic(prep, by = "period"), 12)
```

Making this bias visible is the point: it is exactly the information a functional
prey grouping would smooth over.
