---
title: "Resolving prey taxonomy with WoRMS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Resolving prey taxonomy with WoRMS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

DietView can attach a full taxonomic lineage to your prey by querying the
[World Register of Marine Species (WoRMS)](https://www.marinespecies.org/). This
article explains how the matching works, what provenance it records, and how to
review and correct it. All chunks here are shown but not run (they need the
`worrms` package and internet).

## Why a lineage, and why NA matters

DietView renders diet across the seven flagship Linnaean ranks
(kingdom … species). Your records rarely carry all seven, and identification
often stops at a coarse rank (e.g. `Decapoda`, `Teleostei`). DietView never
invents the missing ranks: where identification stops, the deeper ranks stay
`NA` and are shown as an explicit "NA" category so within-rank percentages still
sum to 100 %.

## The matching strategy

`dv_prepare(taxonomy = "worms")` calls `dv_taxonomy_worms()` on the unique prey.
Matching proceeds in a strict order and every prey carries a provenance record:

1. **id** - the `prey_id` (AphiaID) is already known and used directly;
2. **exact** - the `prey_name` matches a WoRMS latin name exactly;
3. **fuzzy** - if `fuzzy = TRUE`, the closest WoRMS name is proposed (and
   flagged);
4. **unresolved** - nothing matched; all ranks stay `NA`.

Unaccepted names are followed to their **accepted (valid) AphiaID** before the
lineage is fetched, and results are cached to disk so re-runs and other surveys
do not requery.

```{r prepare}
library(DietView)

dat <- read.csv(
  system.file("extdata", "dietview_example.csv", package = "DietView"),
  na.strings = c("NA", "")
)
spec <- dv_spec(
  predator  = "predator_species_latin_name",
  stomach   = "stomach_id",
  prey_id   = "aphia_id",
  prey_name = "verified_name",
  weight    = "prey_weight"
)

prep <- dv_prepare(dat, spec, taxonomy = "worms")   # id / exact / fuzzy / unresolved
prep
```

## Reviewing fuzzy and unresolved matches

Nothing is guessed silently. `dv_taxonomy_report()` returns exactly the prey that
need a human eye, sorted by how many records they affect.

```{r report}
review <- dv_taxonomy_report(prep, which = "review")   # fuzzy + unresolved
review

# write it out for a co-author to check
dv_taxonomy_report(prep, which = "all", file = "taxonomy_log.csv")
```

## Correcting what is wrong

Once you have reviewed the report, fix the source names, or freeze a curated
lineage table and switch to the offline, reproducible path:

```{r fix}
lineage <- read.csv("my_curated_lineage.csv", na.strings = c("NA", ""))
prep <- dv_prepare(dat, spec, taxonomy = "table", lineage = lineage)
```

## Exact-only, conservative matching

To disable fuzzy matching entirely (exact names and known AphiaIDs only):

```{r conservative}
prep <- dv_prepare(dat, spec, taxonomy = "worms", fuzzy = FALSE)
```
