---
title: "Getting started with DietView"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with DietView}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

DietView takes a **plain long table of stomach-content records** — one row per
prey record — and produces taxonomically complete, covariate-aware sunburst
views of predator diet. This article walks the whole pipeline on the bundled
example, fully offline (no internet, no `worrms`), using a lineage lookup table
to attach taxonomy.

```{r setup}
library(DietView)
```

## 1. The input contract

One row is **one prey record in one stomach**. Column names are free — you map
them to DietView's roles once. The bundled base table carries no
`kingdom .. species` columns; DietView adds them.

```{r load}
dat <- read.csv(
  system.file("extdata", "dietview_example.csv", package = "DietView"),
  na.strings = c("NA", "")
)
str(dat)
```

## 2. Declare your columns with `dv_spec()`

Nothing is hard-coded. You describe the roles once and pass the resulting spec to
every other function.

```{r spec}
spec <- dv_spec(
  predator       = "predator_species_common_name",
  predator_latin = "predator_species_latin_name",
  stomach        = "stomach_id",
  prey_id        = "aphia_id",
  prey_name      = "verified_name",
  weight         = "prey_weight",
  digestion      = "digestion_level_id",
  covariates     = c("period", "size_class", "size_bin", "region")
)
```

`dv_validate()` checks that every declared column exists and tells you which
currencies and diagnostics will be available.

```{r validate}
invisible(dv_validate(dat, spec))
```

## 3. Attach taxonomy

Two paths lead to the same prepared object. **Offline / reproducible**: join a
lineage lookup you supply.

```{r prepare-table}
lin <- read.csv(
  system.file("extdata", "dietview_example_lineage.csv", package = "DietView"),
  na.strings = c("NA", "")
)
prep <- dv_prepare(dat, spec, taxonomy = "table", lineage = lin)
prep
```

**Live from WoRMS** (needs the `worrms` package and internet; results are cached
to disk). Fuzzy matches and unresolved prey are flagged, never silent — review
them with `dv_taxonomy_report()`.

```{r prepare-worms, eval = FALSE}
prep <- dv_prepare(dat, spec, taxonomy = "worms")
dv_taxonomy_report(prep)          # audit fuzzy + unresolved matches
```

## 4. Within-rank importance

`dv_build_tree()` aggregates the records into a multi-rank tree. Each rank sums
to 100 %, the explicit `NA` category included, in both occurrence and weight.

```{r tree}
tree <- dv_build_tree(prep$data, prep$ranks, weight_col = spec$weight)
dv_rank_table(tree, "Class")
```

The within-rank shares are a genuine partition:

```{r check-sum}
round(sum(dv_rank_table(tree, "Class")$pct_occ), 3)  # 100
```

## 5. Sunbursts

The multi-rank sunburst reads coarse (centre) to fine (edge). Orientation is
fixed so covariate cells stay directly comparable; unidentified prey (`NA`) is
white.

```{r sunburst-occ}
dv_sunburst(tree, mode = "occurrence", title = "All predators",
            subtitle = "occurrence share within each rank")
```

The same tree in the weight currency:

```{r sunburst-w}
dv_sunburst(tree, mode = "weight", title = "All predators",
            subtitle = "weight share within each rank")
```

## 6. The digestion identification bias

As digestion advances, the finest resolved rank shifts toward coarse ranks and
"Unresolved". `dv_digestion_diagnostic()` makes that bias explicit, optionally
split by a covariate.

```{r digestion}
head(dv_digestion_diagnostic(prep, by = "period"), 10)
```

## 7. Covariate helpers

Derive a period from year, or per-group length bins, and feed them back as
covariates.

```{r covariates}
dv_add_period(c(2004, 2009, 2018), breaks = c(2003, 2007, 2020))
dv_size_bins(c(18, 32, 47), breaks = c(0, 20, 30, 40, 60))
```

## 8. Ship it

A self-contained HTML dashboard (searchable predator cards + one tab per
predator), or the interactive Shiny explorer.

```{r ship, eval = FALSE}
dv_deploy(prep, file = "dietview_dashboard.html", dir = "outputs",
          strata = c("period", "size_class", "size_bin"),
          cross  = list(c("size_bin", "period")),
          length_col = "somatic_length_cm")

run_dietview(prep)
```

See [Building a self-contained diet
dashboard](dashboard.html) for the dashboard options in detail, and [Resolving
prey taxonomy with WoRMS](taxonomy-worms.html) for the online taxonomy path.
