---
title: "Exploring and aggregating diet data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring and aggregating diet data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.5,
  warning = FALSE, message = FALSE
)
```

Before a diet is displayed it is explored, and before it is explored the
reporting resolution has to be chosen. This article covers the functions that
serve those two steps: a small set of descriptive explorers, and a
resolution-robust aggregator. Everything here runs offline on the bundled
example data.

```{r setup}
library(DietView)

csv <- system.file("extdata", "dietview_example.csv", package = "DietView")
lin <- system.file("extdata", "dietview_example_lineage.csv", package = "DietView")
dat <- read.csv(csv, na.strings = c("NA", ""))
lineage <- read.csv(lin, na.strings = c("NA", ""))

spec <- dv_spec(
  predator   = "predator_species_common_name",
  stomach    = "stomach_id",
  prey_id    = "aphia_id",
  prey_name  = "verified_name",
  weight     = "prey_weight",
  count      = "number_of_prey",
  digestion  = "digestion_level_id",
  covariates = c("period", "size_class", "region")
)
prep <- dv_prepare(dat, spec, taxonomy = "table", lineage = lineage,
                   verbose = FALSE)
```

## 1. Describe the sample first

`dv_summary()` answers the questions you ask before anything else: how many
stomachs and records each predator contributed, how many distinct prey taxa
were seen, and the mean prey records per stomach. Pass `by` to break it down by
a covariate.

```{r summary}
dv_summary(prep)
dv_summary(prep, by = "region")
```

A long prey-record table holds only non-empty stomachs, so there is no
vacuity (percent-empty) column -- that has to come from the full sampling frame
upstream.

## 2. Is the diet adequately sampled?

`dv_prey_accumulation()` builds the cumulative prey curve: prey richness against
the number of stomachs, averaged over random stomach orderings. A curve that
flattens means more stomachs would reveal little new; a still-rising curve is a
warning.

```{r accumulation}
acc <- dv_prey_accumulation(prep, permutations = 50, seed = 1)
acc

plot(acc$n_stomachs, acc$mean_richness, type = "b",
     xlab = "Stomachs examined", ylab = "Cumulative prey taxa",
     main = "Prey accumulation (pooled)")
```

Pass `by = spec$predator` for one curve per predator. Keep `permutations`
modest while exploring; raise it for a figure.

## 3. Frequency of occurrence, the classical way

`dv_build_tree()` gives `pct_occ`, a share of prey *records* that is additive so
it can be drawn as a sunburst. That is **not** the classical frequency of
occurrence. `dv_fo_table()` computes the real percent-F -- the share of
non-empty stomachs containing each taxon -- independently at each rank, which is
the nested-hierarchy layout recommended for cross-study comparison
(Buckland et al., 2017).

```{r fo}
dv_fo_table(prep, rank = "class")
```

These values are non-additive across ranks on purpose: a parent's percent-F is
recomputed from stomach presence, not summed from its children. Leave `rank =
NULL` to stack every rank, and pass `by` for a covariate breakdown.

## 4. Resolution-robust aggregation

Keeping the finest rank with explicit NA is the default, but a diet is only
comparable across studies if the reporting resolution is a stated, reproducible
choice rather than an analyst's habit. `dv_aggregate_threshold()` collapses prey
seen in fewer than `threshold` stomachs into their parent, walking up the
hierarchy until every retained category clears the threshold.

```{r aggregate}
agg <- dv_aggregate_threshold(prep, threshold = 3)
table(agg$data$prey_rank, useNA = "ifany")
```

The returned object is a pruned `dietview_prep`: ranks finer than the settled
rank are set to NA, so it drops straight back into the normal pipeline, and the
sunburst then has no leaf rarer than the threshold.

```{r aggregate-tree}
tree <- dv_build_tree(agg$data, agg$ranks, weight_col = spec$weight)
dv_rank_table(tree, "Class")
```

Three architectures, two arguments:

```{r architectures, eval = FALSE}
dv_aggregate_threshold(prep, 100)                      # pooled (primary)
dv_aggregate_threshold(prep, 100, min_predators = 2)   # pooled cross-predator
dv_aggregate_threshold(prep, 50, scope = "per_predator")
```

To reproduce the "resolutions" bookkeeping -- how category count responds to the
threshold -- sweep it:

```{r sweep}
dv_threshold_sweep(prep, thresholds = c(2, 3, 5, 8))
# the paper's grids: seq(10, 1000, 10) -> 100 resolutions;
#                    seq(5, 250, 5) per_predator -> 50 resolutions
```

Reporting a statistic as a mean across resolutions is done in the analysis that
consumes these; DietView only produces the aggregation, not the test.

## 5. Handoff to multivariate analysis

`dv_matrix()` reshapes records into the sampling-unit by prey-category matrix
that ordination, PERMANOVA and compositional regression expect. If you
aggregated first, it uses the `prey_category` column automatically.

```{r matrix}
m <- dv_matrix(agg, currency = "occurrence")
dim(m)
m[1:3, 1:min(5, ncol(m))]
```

`currency` is `"weight"` (summed mass), `"occurrence"` (record counts) or
`"fo"` (presence per unit). Use `unit` to aggregate to something coarser than a
stomach, such as a trawl set, and `include_na = FALSE` to drop the unidentified
column for a clean analysis matrix.

```{r downstream, eval = FALSE}
# DietView stops here; the analysis lives in your own tooling
mm <- vegan::decostand(m, "total")
d  <- vegan::vegdist(mm, "bray")
# vegan::adonis2(d ~ period, data = set_covariates)
```

## Where this fits

These functions cover the exploration and aggregation layers. Inference --
diversity tests, PERMANOVA, dispersion, classification of change -- is
deliberately left to the packages built for it; `dv_matrix()` is the bridge to
them.
