---
title: "A guided tour: every flagship function, what it does, and why"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A guided tour: every flagship function, what it does, and why}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

This article walks the whole package, function by function. For each one it gives
the same four things:

- **Why** the function exists -- the problem it solves;
- **What it does** -- the mechanics;
- **What you get back** -- the actual object;
- **Nuances** -- the parameter choices that change the answer, and the traps.

It closes with side-by-side applications that show how much the answer moves when
you change a currency or a reporting rank. Everything runs offline on the bundled
example data.

```{r setup}
library(DietView)
```

## The mental model

DietView is a short linear pipeline carrying three objects. Nothing is hidden:
every stage returns something you can print, inspect and export.

| Stage | Function | Object produced |
|---|---|---|
| Declare | `dv_spec()` | `dietview_spec` -- the column mapping |
| Check | `dv_validate()` | nothing; reports on the data |
| Prepare | `dv_prepare()` | `dietview_prep` -- data + taxonomy + provenance log |
| Review | `dv_taxonomy_report()` | data.frame of matches needing a human eye |
| Aggregate | `dv_build_tree()` | `dietview_tree` -- nodes with within-rank importance |
| Read | `dv_rank_table()` | data.frame -- the numbers behind one ring |
| Show | `dv_sunburst()` | plotly figure |
| Ship | `dv_build_html()`, `run_dietview()`, `dv_deploy()` | dashboard, app, deployment |

Two helpers (`dv_add_period()`, `dv_size_bins()`) build covariates, one
(`dv_palette()`) controls colour, one (`dv_digestion_diagnostic()`) quantifies
identification bias, and one (`dv_species_card_assets()`) fetches identification
material.

```{r load}
dat <- read.csv(
  system.file("extdata", "dietview_example.csv", package = "DietView"),
  na.strings = c("NA", "")
)
lineage <- read.csv(
  system.file("extdata", "dietview_example_lineage.csv", package = "DietView"),
  na.strings = c("NA", "")
)

str(dat, vec.len = 2, give.attr = FALSE)
```

One row is **one prey record in one stomach**. There are no `kingdom .. species`
columns -- DietView adds them.

## `dv_spec()` -- declare your columns once

**Why.** Every survey names its columns differently. The alternative to a
specification object is renaming columns to canonical names before analysis,
which is a destructive, undocumented step performed outside the analysis: the
published code then operates on a table that no longer matches the archived data.
A spec keeps the original table untouched and makes the mapping an artefact you
can print, archive and compare between studies.

**What it does.** Stores which column plays which role. Nothing else. No data is
touched.

```{r spec}
spec <- dv_spec(
  predator   = "predator_species_common_name",
  stomach    = "stomach_id",
  prey_id    = "aphia_id",
  prey_name  = "verified_name",
  weight     = "prey_weight",
  count      = "number_of_prey",
  digestion  = "digestion_level_id",
  covariates = c("period", "size_class", "region")
)

str(spec)
```

**Nuances.**

*`prey_id` takes precedence over `prey_name`.* Supply both when you have both:
the identifier resolves unambiguously, and the name is kept for the provenance
log so you can read what you are reviewing. Where the identifier is missing, the
name is the fallback.

*`prey_name` must be the scientific (Latin) name*, because it is sent to WoRMS,
which does not know common names. A common name either fails to resolve or --
worse -- fuzzy-matches to a plausible wrong lineage. Note the asymmetry with
`predator`, which is a **common** name here: the predator is never resolved
against any authority, it is only a grouping and display label.

*The name does not have to be a species.* `Decapoda`, `Amphipoda` and `Teleostei`
are all valid values -- they are simply the rank at which identification stopped.
Do not invent a species to "complete" a record; that is exactly what the explicit
NA machinery is for.

*`ranks` defaults to the seven Linnaean ranks* (`kingdom` to `species`). Shorten
it if your prey are never resolved past family; the sunburst then has fewer rings
and reads more clearly.

*`covariates` is what you can later stratify and cross by.* Declaring a column
here costs nothing and enables it downstream, so declare generously.

## `dv_validate()` -- see the shape of your data before you trust it

**Why.** Most diet datasets have problems that stay invisible until they have
already distorted a figure. Validation surfaces them at the point where you can
still act.

**What it does.** Confirms declared columns exist, that a prey key is present,
that covariates exist, and reports on what is missing. It **reports rather than
repairs**: a dataset with a third of weights missing is analysable in the
occurrence currency and misleading in the weight currency, and that judgement is
yours to make.

```{r validate}
dv_validate(dat, spec)
```

**What you get back.** The data, invisibly -- so the value is in the messages and
warnings, not the return value. Errors are hard stops (a declared column absent);
warnings and messages are informational (no weight column, no digestion column).

**Nuance.** Silence is a pass. If you see nothing, every declared column was
found and the currencies you asked for are available.

## `dv_prepare()` -- attach taxonomy, and log how

**Why.** This is where grouping stops being an analyst's judgement call and
becomes a lookup against an external, versioned authority. It is the single
change that makes two surveys comparable without recoding.

**What it does.** Validates, resolves the **distinct** prey keys against a
taxonomic source, joins the lineage back to every record, and fills the ranks
below the identification point with an explicit `"NA"` marker.

```{r prepare}
prep <- dv_prepare(
  dat, spec,
  taxonomy = "table",   # offline: use the supplied lineage
  lineage  = lineage,
  verbose  = FALSE
)

class(prep)
names(prep)
```

**What you get back.** A `dietview_prep` list with four components:

- `data` -- your original table plus the rank columns;
- `spec` -- the spec, carried along;
- `ranks` -- the ranks actually populated;
- `taxonomy_log` -- one row per distinct prey, with how it resolved.

```{r prep-ranks}
prep$ranks
head(prep$data[, c("verified_name", prep$ranks)], 4)
```

**Nuances.**

*`taxonomy =` chooses the source and matters more than any other argument here.*

| Value | Behaviour | When to use |
|---|---|---|
| `"auto"` | use rank columns if present, else WoRMS | quick exploration |
| `"columns"` | trust rank columns already in the data | taxonomy resolved elsewhere |
| `"worms"` | query WoRMS online | first pass on a new dataset |
| `"table"` | join a supplied lineage table | reproducible reruns, offline work, non-marine prey |

*The recommended production pattern is resolve once, then freeze.* Run with
`taxonomy = "worms"` a single time, review the log, export the resulting lineage,
and thereafter run with `taxonomy = "table"`. This makes later runs offline and
fast, and it pins the taxonomic snapshot -- different versions of an authority
can return different lineages, so an archived lineage is to taxonomy what a
lockfile is to package versions.

*`fuzzy = TRUE` is a convenience with a cost.* It rescues misspellings, and it can
silently attach a plausible wrong lineage. It is safe only because every fuzzy hit
is logged -- which is worth nothing if you do not read the log. See the next
function.

*Resolution scales with taxonomic diversity, not with sample size.* Two million
records containing three thousand distinct prey cost three thousand lookups.

## `dv_taxonomy_report()` -- read the log before you plot

**Why.** Fuzzy matches and unresolved names are the two places where a diet
description silently goes wrong. This is the review step.

**What it does.** Returns the rows of the provenance log that need a human eye.

```{r taxreport, eval = FALSE}
prep_online <- dv_prepare(dat, spec, taxonomy = "worms")

dv_taxonomy_report(prep_online, which = "review")   # fuzzy + unresolved
dv_taxonomy_report(prep_online, which = "all",      # archive the whole log
                   file = "taxonomy_log.csv")
```

**Nuance you will meet immediately.** The provenance log only exists on the WoRMS
path. A `prep` built with `taxonomy = "table"` -- as everywhere else in this
article, so that it runs offline -- has `taxonomy_log = NULL`, and this function
stops with a message saying so. That is deliberate: a lineage table you supplied
yourself has no matching history to report, because you did the matching. It also
means the review step belongs to the **one** online run in the
resolve-once-then-freeze pattern, not to every rerun.

`which = "review"` (the default) is fuzzy plus unresolved -- the actionable set.
`"fuzzy"` and `"unresolved"` isolate one or the other; `"all"` returns the full
log, which is what you archive alongside the analysis. Pass `file =` to write it
out.

Unresolved is not necessarily an error. In this dataset `"unidentified material"`
has no AphiaID and no taxonomic meaning, so it *should* fail to resolve -- it
becomes the explicit NA category. Unresolved entries you did not expect are the
ones to chase.

## `dv_build_tree()` -- the metrics core

**Why.** This is where within-rank importance is computed, and it is the
analytical heart of the package.

**What it does.** Aggregates at the finest rank, then rolls up. Three quantities
travel with every node: `value_item` (distinct terminal items), `value_occ`
(occurrence count), `value_w` (summed weight). It then divides each by its
**rank total** to give `pct_occ` and `pct_w`.

```{r tree}
tree <- dv_build_tree(
  prep$data,
  ranks      = prep$ranks,
  weight_col = spec$weight
)

head(tree, 4)
```

**What you get back.** A `dietview_tree` data.frame: `id`, `label`, `parent`,
`rank`, the three values, and the two within-rank percentages.

Two properties hold **by construction**, and both are worth verifying once so you
trust them thereafter:

```{r invariants}
# 1. every rank is a complete partition (each sums to 100%)
round(tapply(tree$pct_occ, tree$rank, sum), 6)

# 2. a parent's value equals the sum of its children (largest gap should be 0)
kid_sums <- tapply(tree$value_occ, tree$parent, sum)
parents  <- intersect(tree$id, names(kid_sums))
max(abs(tree$value_occ[match(parents, tree$id)] - kid_sums[parents]))
```

The first is what lets you read any ring on its own. The second is what a nested
sunburst requires -- `branchvalues = "total"` draws a parent as the sum of its
children, and a tree that violates this renders inconsistently.

**Nuances.**

*`weight_col = NULL` is legal* and gives an empty weight view. Occurrence still
works. This is the right setting when weights are too sparse to trust.

*`count_col` changes what one "occurrence" means.* Left `NULL`, each record
contributes 1. Supplied, each record contributes its count -- which turns the
occurrence currency into a numerical (%N-like) currency. Choose deliberately and
say which you used.

*`value_occ` is a count of prey records, not a count of stomachs.* This is the
one thing about the package most likely to be misread. It is a share of records,
close in spirit to %N; it is **not** the frequency of occurrence (%F) of the
classical literature, which counts stomachs and whose values at a parent rank are
*not* the sum of its children (a stomach containing two congeners counts once for
the genus but twice among the species). A genuine %F cannot be drawn as a nested
sunburst for exactly that reason: the geometry demands additivity and %F is not
additive. Report the currency by its real name.

## `dv_rank_table()` -- the numbers behind one ring

**Why.** A figure is for seeing; a table is for reporting. Journals want numbers,
and so does anyone checking your work.

**What it does.** Extracts one rank from the tree, sorted by occurrence share,
with the explicit NA row included.

```{r ranktable}
dv_rank_table(tree, "Class")
```

**Nuances.** Rank labels are title-case (`"Class"`, `"Family"`, `"Genus"`) --
they come from the spec's `ranks`, capitalised. The NA row is deliberately kept:
dropping it is exactly the silent re-basing this package exists to prevent.

## `dv_sunburst()` -- the whole diet in one figure

**Why.** Taxonomy is hierarchical; bar charts are flat. One sunburst replaces a
wall of per-rank, per-currency panels.

**What it does.** Renders the multi-rank tree, unidentified prey in white.

```{r sunburst-occ}
dv_sunburst(tree, mode = "occurrence", title = "Pooled diet, occurrence")
```

**Nuances -- the three arguments that matter.**

*`mode` selects the currency.* `"occurrence"`, `"weight"`, `"count"`. See the
application below: the first two routinely disagree, and showing both is part of
an honest description.

*`colors` is the one you must not forget when comparing panels.* Left `NULL`,
each call derives its own palette from **its own** labels -- so a panel with 12
taxa and a panel with 9 give the same taxon different colours, and the comparison
you were trying to make is destroyed. Build one global palette and pass it
everywhere:

```{r global-palette}
colors <- dv_palette(tree$label)   # from the pooled tree: covers every taxon
head(colors, 4)
colors["NA"]                       # unidentified is always white
```

`run_dietview()` and `dv_build_html()` already do this internally; it is manual
per-panel plotting that needs the discipline.

*`rotation = 90` fixes the start angle at the top*, and internally `sort = FALSE`
stops plotly reordering wedges by size. Together with the alphabetical node order
produced by `dv_build_tree()`, this keeps reading order and start angle constant
across panels.

One honest limit: this fixes reading order and start angle, **not** absolute
angular position. A panel missing a taxon shifts the wedges that follow it, since
its tree contains only the taxa present in that stratum. Colour, not angle, is the
reliable cross-panel anchor.

## `dv_palette()` -- colour, and the white NA

**Why.** Comparability across panels is carried by colour, and the unidentified
category needs to be unmistakable.

**What it does.** Maps labels to a categorical palette and forces `"NA"` to white.

**Nuance.** Always derive it from the **pooled** tree, never from a stratum --
the pooled tree is the only one guaranteed to contain every taxon.

## `dv_add_period()` and `dv_size_bins()` -- covariates in one line

**Why.** Size structures diet more strongly than almost anything else, because
gape limitation determines what is available at all. A diet description averaged
over a predator's whole size range is averaging over an ontogenetic trajectory.
Making stratification cheap is what makes it routine.

```{r covariates}
dat$period2   <- dv_add_period(dat$year, breaks = c(1990, 2000, 2010, 2025))
dat$size_bin2 <- dv_size_bins(dat$somatic_length_cm, breaks = c(0, 30, 50, 200))

table(dat$period2, useNA = "ifany")
table(dat$size_bin2, useNA = "ifany")
```

**Nuances.** `dv_add_period()` labels intervals `"1990-2000"` and so on unless
you pass `labels`. `dv_size_bins()` uses `right = FALSE` by default, so bins are
left-closed (`[0-30[`) -- a fish of exactly 30 cm falls in the *second* bin.
Choose breaks on biology (maturity, gape) rather than on quantiles, and state
them in the methods.

## `dv_digestion_diagnostic()` -- make the bias visible

**Why.** Digestion determines both what can be identified and what remains to be
weighed. Reporting that relationship is what turns a nuisance variable into a
diagnostic -- and it is what the standardisation literature has asked studies to
provide.

**What it does.** Cross-tabulates digestion level against the finest rank
resolved, in records and in weight.

```{r digestion}
dg <- dv_digestion_diagnostic(prep)
head(dg, 8)
```

**What you get back.** A data.frame with `digestion`, `finest_rank`, `n_records`,
`weight`, and the within-digestion-level percentages `p_records` and `p_weight`.

**Nuances.** Percentages are computed **within** each digestion level, so each
level sums to 100 and levels with few records are not visually swamped. Pass
`by =` to add a third dimension (predator, region) when you suspect the
degradation profile differs between strata. A steep profile is a warning that the
weight currency in particular deserves caution.

## `dv_taxonomy_worms()` -- the resolver, on its own

**Why.** `dv_prepare(taxonomy = "worms")` calls this internally. Calling it
directly is what you do to build the lineage table you will freeze.

**What it does.** Takes a data.frame of unique prey keys and returns their
lineages plus the match log. It tries, in order: retrieval by AphiaID, the
accepted ID behind an unaccepted name, exact name match, then fuzzy match.

```{r worms, eval = FALSE}
keys <- unique(dat[, c("aphia_id", "verified_name")])
names(keys) <- c("prey_id", "prey_name")

tax <- dv_taxonomy_worms(keys, fuzzy = TRUE)

table(tax$match_type)              # id / exact / fuzzy / unresolved
write.csv(tax, "my_lineage.csv", row.names = FALSE)   # freeze it
```

**Nuances.** It expects columns named `prey_id` and `prey_name`, one row per
unique prey -- not your full record table. Nothing is guessed silently: every row
carries its `match_type` and `matched_name` so a fuzzy hit can be corrected by
hand before you save the lineage. Results are cached under `cache_dir`, so a
rerun after a corrected spelling only queries what changed.

## `dv_species_card_assets()` -- identification material

**Why.** The audience for a dashboard often needs to *recognise* the taxa, not
just read their proportions: technicians calibrating identification, students on
an unfamiliar fauna, managers new to the system.

```{r cards, eval = FALSE}
dv_species_card_assets(
  c("Pandalus borealis", "Osmerus mordax"),
  use_wiki  = TRUE,
  image_dir = tempdir()
)
```

**Nuances.** `use_wiki = FALSE` builds cards offline -- taxonomic information
only, no imagery. Images are fetched from an open source and written to
`image_dir`, so point it somewhere durable if you want the dashboard to keep
working after a reboot.

## `dv_build_html()`, `run_dietview()`, `dv_deploy()` -- three audiences

**Why.** The audience for a diet description frequently does not run R. A result
that needs a live session is a result that will be transcribed into a static
figure and lose its structure.

```{r ship, eval = FALSE}
# self-contained dashboard: one file, opens in any browser, no R
dv_build_html(
  prep,
  output = file.path(tempdir(), "dietview.html"),
  strata = c("period", "size_class"),
  cross  = list(c("size_class", "period"))   # inner, outer
)

# live exploration
run_dietview(prep)
```

**Nuances.** `cross = list(c(inner, outer))` reads as "inner **by** outer": one
row per level of `outer`, one column per level of `inner`. Crossing multiplies
panels, so cost and readability both degrade with the product of the level
counts -- two covariates with four levels each is sixteen sunbursts.

A dashboard is a **snapshot**, not a live view: it embeds its data at build time.
That is a limitation for monitoring and a virtue for archiving.

## Application 1 -- the two currencies disagree

This is the comparison the package exists to make easy. Same tree, same records,
different contribution per record.

```{r app-currencies}
cls <- dv_rank_table(tree, "Class")

data.frame(
  taxon        = cls$label,
  occurrence   = round(cls$pct_occ, 1),
  weight       = round(cls$pct_w, 1),
  difference   = round(cls$pct_w - cls$pct_occ, 1)
)
```

Large, rarely eaten prey rank high by weight and low by occurrence; small,
frequently eaten prey do the reverse. Neither column is the right answer.
Reporting one without the other is a choice that should be stated, not a default.

## Application 2 -- how much does the reporting rank change the story?

Because the tree carries every rank at once, a sensitivity analysis that is
normally awkward becomes three lines. This is the exercise conventional workflows
cannot perform at all, since the finer ranks were discarded before analysis.

```{r app-ranks}
for (r in c("Class", "Order", "Family")) {
  tb <- dv_rank_table(tree, r)
  cat("\n--", r, "-- top 3 by occurrence share:\n")
  print(head(tb[, c("label", "pct_occ")], 3), row.names = FALSE)
}
```

Watch for the classic pattern: **stability at coarse ranks concealing turnover at
fine ranks**. If two strata look alike at class and separate at genus, the
substitution is happening *within* your coarse categories -- which is precisely
what a functional-group analysis would have reported as "no change".

## Application 3 -- how much of the diet was actually identified?

The explicit NA category is not decoration; it is the confidence bound on every
fine-rank statement in the figure.

```{r app-na}
na_share <- do.call(rbind, lapply(prep$ranks, function(r) {
  lab <- paste0(toupper(substring(r, 1, 1)), substring(r, 2))
  tb  <- dv_rank_table(tree, lab)
  data.frame(rank = lab,
             na_occurrence = round(sum(tb$pct_occ[tb$label == "NA"]), 1))
}))
na_share
```

Read this as a ceiling on trust. If half the records are unidentified at genus,
a genus-level claim rests on the identifiable half -- and digestibility is not
random with respect to taxon, so that half is not a random sample.

## What changes what -- a summary

| Argument | Where | Changes |
|---|---|---|
| `prey_id` / `prey_name` | `dv_spec()` | which key resolves; ID wins when both given |
| `ranks` | `dv_spec()` | number of rings in the sunburst |
| `covariates` | `dv_spec()` | what you can stratify and cross by later |
| `taxonomy` | `dv_prepare()` | the source of grouping, and reproducibility |
| `fuzzy` | `dv_prepare()` | recovers misspellings; needs the log reviewed |
| `weight_col` | `dv_build_tree()` | whether the weight currency exists |
| `count_col` | `dv_build_tree()` | occurrence becomes a numerical currency |
| `mode` | `dv_sunburst()` | which currency is shown |
| `colors` | `dv_sunburst()` | cross-panel comparability -- pass a global palette |
| `rotation` | `dv_sunburst()` | start angle; keep constant across panels |
| `breaks` | `dv_add_period()`, `dv_size_bins()` | the strata themselves |
| `by` | `dv_digestion_diagnostic()` | whether the bias is broken down further |
| `cross` | `dv_build_html()` | the panel grid, as inner-by-outer |

## Three things to carry away

**The denominator is stated, per rank.** Every ring sums to 100% including the
explicit NA, so nothing is silently re-based onto the identified fraction. That
is what makes the rings independently readable.

**The occurrence currency is a share of records.** It is not the classical %F,
which counts stomachs and is non-additive. Name it correctly in your methods.

**Reproducibility is a workflow, not a function.** Resolve the taxonomy once,
read the provenance report, archive the lineage table, and run everything
afterwards from that snapshot with the spec. What makes a DietView figure
reproducible is those three artefacts -- the data, the spec and the lineage --
not the plotting call.
