---
title: "Within-rank importance and the explicit-NA philosophy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Within-rank importance and the explicit-NA philosophy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5,
                      warning = FALSE, message = FALSE)
```

This article explains the two ideas at the heart of DietView: importance is
computed **within each taxonomic rank**, and unidentified prey is kept as an
**explicit "NA" category** rather than dropped. Together they make the sunburst an
honest partition of the diet.

```{r setup}
library(DietView)

dat <- read.csv(system.file("extdata", "dietview_example.csv", package = "DietView"),
                na.strings = c("NA", ""))
lin <- read.csv(system.file("extdata", "dietview_example_lineage.csv", package = "DietView"),
                na.strings = c("NA", ""))
spec <- dv_spec(predator = "predator_species_common_name", stomach = "stomach_id",
                prey_id = "aphia_id", prey_name = "verified_name",
                weight = "prey_weight", digestion = "digestion_level_id",
                covariates = c("period", "size_class", "size_bin", "region"))
prep <- dv_prepare(dat, spec, taxonomy = "table", lineage = lin)
```

## Building the tree

`dv_build_tree()` aggregates records into a node table spanning all ranks, in two
currencies (occurrence and weight).

```{r tree}
tree <- dv_build_tree(prep$data, prep$ranks, weight_col = spec$weight)
head(tree[, c("rank", "label", "value_occ", "pct_occ", "value_w", "pct_w")], 8)
```

## Each rank is its own 100 %

Importance is the share of a node **within its own rank**, so every rank sums to
100 % independently — the explicit `NA` node included.

```{r sums}
sapply(unique(tree$rank), function(r) round(sum(tree$pct_occ[tree$rank == r]), 3))
```

`dv_rank_table()` gives the sorted importance table for one rank:

```{r ranktable}
dv_rank_table(tree, "Class")
```

## NA is information, not an error

Where identification stops, DietView keeps an explicit `NA` node (white in the
sunburst) instead of rebasing percentages onto the identified fraction. The `NA`
row is present at every rank where some prey are unresolved:

```{r na}
subset(dv_rank_table(tree, "Order"), label == "NA")
```

## The sunburst

Reading coarse (centre) to fine (edge), with `NA` shown in white:

```{r sunburst-occ}
dv_sunburst(tree, mode = "occurrence", title = "All predators",
            subtitle = "occurrence share within each rank")
```

Occurrence and weight routinely disagree — showing both is part of an honest diet
description:

```{r sunburst-w}
dv_sunburst(tree, mode = "weight", title = "All predators",
            subtitle = "weight share within each rank")
```

## A stable colour dictionary

`dv_palette()` assigns each taxon a stable colour so it keeps the same colour
across ranks and covariate cells; `NA` is always white.

```{r palette}
pal <- dv_palette(tree$label)
head(pal)
pal[["NA"]]
```
